Effects of bariatric surgery and weight loss on resting state functional connectivity

### 2021-01-29 manuscript/draft_rsfmri.Rmd
all computations transferred to process.R instead of .Rmn

---
title: "Effects of bariatric surgery and weight loss on resting state functional connectivity"
author:
- Hannah Sophie Heinrichs* [1]
- Frauke Beyer* [1][2]
- Kerstin Prehn
- Agnes Floel
- Arno Villringer [1][2][3]
- A. Veronica Witte [1][2][3]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3
  Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  bookdown::pdf_book:
    fig_caption: yes
    always_allow_html: true
    number_sections: yes
    toc: yes
    extra_dependencies: "subfig"
bibliography: "bibliography.bib"
editor_options:
  chunk_output_type: console
---

@@

```{r "knit setup", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex")
options(knitr.kable.NA = '') # hide NA in tables
# set working directory of knit (RMarkdown) to directory of Rproj
opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
setwd(knitr::opts_knit$get("root.dir"))

source("code/figures.R")
source("code/tables.R")
source("code/create_sample_df.R")
```

@@

```{r "Load data", include=FALSE}
#final <- create_sample_df(group = "both", tp = "all")
final <- readRDS(file = "./report/tab/final.rds")

@@

```{r "tableSample", echo=FALSE}
#tableSample <- mk_SampleTable(final)
tableSample <- readRDS(file = "report/tab/tableSample.rds")

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study conducted in the Charité Berlin is a observational, prospective and longitudinal design described in detail by @Prehn_2020. Data collection is still ongoing but only data available in April 2019 is used.
The procedures followed were in accordance with the Helsinki Declaration. This study was registered at clinicaltrials.gov as NCT01554228 and described in more detailed in the preregistration under [https://osf.io/yp42s](https://osf.io/yp42s). <!-- previous version: https://osf.io/f8tpn, update: https://osf.io/59bh7/ -->
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA --> Participants were recruited from the Center for Bariatric and Metabolic Surgery if they were between age $18$ and $70$ and fulfilled the inclusion criteria. These were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013].
Participants with a history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$) were excluded from the study.
Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment but not all participants have complete data for all time points (see Table \@ref(tab:tableDescr)). `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point.
The current sample is the subsample of $69$ subjects that were originally enrolled in the study entailing only subjects that have received MRI. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data availability in \@ref(tab:tableSample)).

<!-- DESCRIPTIVE STATISTICS --> The final sample entailed morbidly obese individuals 
(n = `r nrow(subj_df)`, 
`r nrow(filter(subj_df,Sex == "f"))` female; 
`r round(mean(subj_df$Age_BL),2)` 
$\pm$ `r round(sd(subj_df$Age_BL),2)` SD years old
(range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`), 
that either underwent surgery 
(n = `r cs["Sum","IG"]`, 
`r cs["f","IG"]` female, 
`r round(mean(subj_df$Age_BL[subj_df$condition == "IG"]),2)` 
$\pm$ `r round(sd(subj_df$Age_BL[subj_df$condition == "IG"]),2)` SD years old) 
or were waiting list controls 
(n = `r cs["Sum","KG"]`, 
`r cs["f","KG"]` female,
`r round(mean(subj_df$Age_BL[subj_df$condition == "KG"]),2)`
$\pm$ `r round(sd(subj_df$Age_BL[subj_df$condition == "KG"]),2)` SD years old) 
The study design was quasi-experimental, as patient in the waiting control list had also been recommended to undergo bariatric surgery, but were still waiting for their health insurance's approval.
```{r tableBaselineCharacteristics, include=FALSE}
#dfBL <- create_sample_df(group = "all", tp = "BL")
dfBL <- readRDS(file = "./report/tab/dfBL.rds")

@@

# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555")
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```
<!-- BASELINE CHARACTERISTICS --> Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). Change in BMI throughout the study is depicted in \@ref(fig:figBMIdescr).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
```{r tableDescr, echo=FALSE}

# table_descr <- mk_tableDescr()
# table for descriptive statistics

tableDescr <- readRDS(file = "./report/tab/tableDescr.rds")

tab <-
  knitr::kable(
    tableDescr,
    row.names = FALSE,
    format = "latex",
    booktabs = T,
    linesep = "",
    caption = "BMI and mean FD of participants acquired in each condition and time point")

tab

```
<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropometric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).
```{r figBMIdescr, fig.align='center', out.width="90%", fig.cap="BMI over time.", echo=FALSE}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
fig_BMIdescr <- readRDS(file = "./report/fig/fig_BMIdescr.rds")
fig_BMIdescr
```

@@

### Minimal functional preprocessing
The function preprocessing module included the removal of first four volumes, motion correction (FSL's `MCFLIRT`), fieldmap distortion correction (FSL's `fsl_prepare_fieldmap` and `FUGUE`) and coregistration to the subject’s individual longitudinal anatomical space (FreeSurfer's `bbregister`). In more detail, the transformations derived from the latter three steps were combined into one and applied in a single step.
For further analysis and ICA-AROMA processing, the minimally preprocessed data were intensity normalized and smoothed with a 6mm Gaussian kernel (`fslmaths -kernel gauss 2.548`).
*This part of the analysis was described in the preregistration as "After removing the first four volumes, transforms for motion correction (FSL’s MCFLIRT), fieldmap distortion correction (based on fieldmap and FSL’s FUGUE) and coregistration to the subject’s individual anatomical space (using FREESURFER’s bbregister) are calculated."*

@@

### Functional connectivity
To derive reward and DMN functional connectivity maps, we used Nucleus Accumbens (Nacc) and precuneus as seed regions of interest (ROI), respectively. Based on FreeSurfer's `aseg.mgz` and Desikan-Killiany parcellation, we created seed masks using `mri_binarize` (thresholded for NAcc at 26,58; precuneus at 1025,2025) and averaged them over hemispheres. Then, we used `NiftiLabelsMasker` and `NiftiMasker` to extract the standardized timeseries from the seed regions and the whole brain. We calculated the Pearson's correlation between them with `numpy.dot`, performed r-to-z transform and saved the resulting correlation maps for each processing mode (minimally preprocessed, aroma+cc, aroma+gsr). Finally, the connectivity maps were transformed into MNI space using the affine transformation and non-linear warp derived with ANTS during anatomical preprocessing.
*In the preregistration we stated "Resting-state functional connectivity (FC) using seed-based connectivity analysis extracted with in-house pre-processing pipelines (see (Beyer et al., 2019)): Whole-brain FC maps for a priori defined seed regions (defined based on the results of the subcortical segmentation and Desikan-Killiany cortical parcellation in the Freesurfer version 6.0.0 longitudinal processing stream). For the reward network a ROI combined from bilateral Nucleus accumbens from the subcortical segmentation (26, Left-Accumbens-area, 58, Right-Accumbens-area) was selected. Ventro-medialprefrontal cortex was not selected as seed region due to low signal-to-noise ratio in this part of the brain in many participants (noticed during quality checking). For the default mode network, the seed region was combined from bilateral precuneus from the Desikan-Killiany atlas (1025 ctx-lh-precuneus, 2025 ctx-rh-precuneus). We will calculate  voxel-wise seed-based connectivity mapsin the individual subject’s space (which is coregistered to the Freesurfer longitudinally preprocessed timepoint) and apply r-to-Z-transform to these images. These steps will be performed using nilearn, nipype and python. Then, we will transform the connectivity maps into MNI space with a non-linear warp derived from anatomical preprocessing of the longitudinally preprocessed timepoints (using ANTS).*

@@

FreeSurfer cross-sectional and longitudinal  segmentations were visually checked according to @Klapwijk_2019. We excluded 5 datasets from 3 participants because of excessive head motion leading to failed pial reconstruction and anatomical-functional coregistration .
<!--ADI009_fu2, ADI063_bl, ADI063_fuADI116_bl, ADI116_fu, ADI116_fu2-->
*In the preregistration we stated that "Freesurfer segmentations and surfaces will be visually controlled and corrected according to published recommendations, if necessary", and that "We will exclude participants in which head motion artifacts affected the T1-weighted image preprocessing (e.g. caused Freesurfer to fail or perform very poorly (rated “Failed” according to (Klapwijk et al., 2019)".*

@@

<!-- param.ROI_PREP = {roi_prep{[6,14]}}; % {roi_prep{[4, 6, 12, 14, 20, 22, 28, 30]}} or {'Nacc_cc_z','Nacc_gsr_z','PCC_cc_z','PCC_gsr_z','LH_cc_z','LH_gsr_z','MH_cc_z','MH_gsr_z'} -->
<!-- % Model definition -->
<!-- % All three models have unique options for covariate definition, the -->
<!-- % association to a model is indicated by the tens digit (GroupTime_: 1_;  -->
<!-- % BMI_ = 2_; FD_ = 3_) the specific covariate combination by the ones digit -->
<!-- param.MODEL = {'alltp'}; % {'grouptime','grouptime2tp'} % {'bmi','bmiIG','bmi2tp'} % {'fd','fdIG'} % {'alltp'} % {'singletp}  -->
<!-- param.COVARIATES = [41];     % [11, 12];                    % [21, 22];                % [31, 32];  % [41, 42]    % [41, 42, 43] -->

<!-- % define masking and type of inference -->
<!-- param.MASK = 'brain';               % 'brain' -->
<!-- param.EXCLFD = false;               % false -->
<!-- param.WILD_BOOT = false;            % false -->
<!-- param.INFERENCE_TYPE = {'cluster'}; % {'voxel','cluster','tfce'}; -->

*In the preregistration we wrote: "We will use the average of connectivity values within masks of the respective networks as aggregate network connectivity measure. The mask will include all voxels surviving FWE-correction of p<0.05 in a one-sample t-test of the individual connectivity maps across all participants and time points, calculated with the modified SwE option in the SwE toolbox"*

<!-- MOTION CORRECTION We therefore apply best-practice motion correction (ICA-AROMA and CC, plus optionally GSR) and monitor the reduction of motion-related artifact according to current recommendations [@Ciric_2018]. As quality checks statistical analysis were performed on both with and without global signal regression, considering the controversial discussion about GSR introducing negative correlations and spatial bias into connectivity data [@Ciric_2017; @Murphy_2017].
Additionally, mean framewise displacement (FD) was extracted from 6 translational/rotational motion parameters of the whole time series according to @Power_2012, log-transformed and examined within and separate from the main models.  -->

@@

DesignMatricesList <- readRDS(file = "./report/fig/DesignMatricesList.rds")
DesignMatricesList$plotA
DesignMatricesList$plotB
#gridExtra::grid.arrange(plotA, plotB, nrow=1)
```
For the second and third model, we calculated the centered mean BMI and log mean FD per subject and the intra-subject-centered BMI as proposed by @Guillaume_2014. With a model containing average BMI and longitudinal change per subject, we investigated their differential effects on the complete sample and subsequently in a exploratory fashion for the intervention group only as the effect of longitudinal change in BMI was expected to be driven by this subsample.
<!-- NUISANCE COVARIATES --> Baseline age, sex, and log mean framewise displacement (FD) were entered into the first and second model as nuisance variables. Age and sex were between-subject covariates while FD is time-varying. Mean FD was log-transformed and age was demeaned by subtracting the overall mean across subjects and time points.
As change in FD and change in BMI share common variance [@Beyer_2019], the inclusion of log mean FD may be disputed and done optionally. For that reason, we performed each analysis separately one with mean FD (three and without. Differences in results will be reported. We consider this sensitivity analysis theoretically more meaningful than the preregistered comparison with the nuisance covariate free model and a model with four nuisance covariates (including baseline BMI).<!-- Prereg: We will perform sensitivity analysis adjusting for baseline age, sex, average of mean FD of both time points and baseline BMI -->
<!-- POST HOC  --> For a better understanding of the unique contribution of average and longitudinal change FD measures, we separately employed a third post-hoc model to disentangle the effects: $FC = between-subject FD + within-subject FD$ with nuisance covariates age and sex. As these post-hoc analyses were not preregistered, they should be considered to be exploratory.

@@

## Quality control
We performed different quality controls for the rsfMRI data.
As described in @Beyer_2020, group and time show an interaction effect for mFD where the intervention group reduced in head motion in the follow-up assessments compared to the control group.
We also inspected DVARS, a measure of whole-brain signal variation which is reflective of noise in the data. We saw that DVARS did not qualitatively differ between groups and time points, and that the correlation of DVARS and mFD remained similar over time points.
This indicates that even though there was a decrease in head motion in the intervention versus control group, the impact of head motion on imaging data remained similar over the study period.
```{r tableQcfc, echo=FALSE}
res <- read.csv("./report/FDFCcorrelations.csv")
tab <- readRDS(file = "./report/fig/tableQcfc.rds")
tab
```
Further, we evaluated the impact of our denoising pipelines on different quality metrics following [@Ciric_2018]. Table \@ref(tab:tableQcfc) shows that in the minimally preprocessed data, mFD and FC between nodes was significantly positively correlated (mean Spearman's r = `r round(res[1,1],2)`). This relationship was negatively associated with distance between nodes (mean Spearman's r = `r round(res[1,5],2)`).  
Denoising procedures (AROMA, AROMA+CC, AROMA+CC+GSR) gradually reduced the correlation of mFD and connectivity, and reduced the distance dependency. Contrary to previous reports, GSR did not exercabate mFD-FC distance dependency. Overall, AROMA+CC+GSR performed best, with almost no mFD-FC correlation (mean Spearman's r = `r round(res[4,1],2)`), and minimal positive distance dependence (mean Spearman's r = `r round(res[4,5],2)`). Yet, AROMA+CC also performed well, and we thus calculated our analysis models for both preprocessing schemes.

```{r figDvarsmFD, fig.align = "center", out.width="100%", fig.cap='Summary of quality measures DVARS and mFD', echo=FALSE}
final=read.csv(file = "report/all_QA_measures.csv")

FigureList <- figDvarsmFD(final)
(FigureList$fig1 | FigureList$fig2) / FigureList$fig3
```

```{r FDFC correlation, eval=FALSE, echo=FALSE}
# do not evaluate if saved already
source("/data/pt_02161/Analysis/Preprocessing/qa/rs_qa/group_level_QA/QC_FC_correlations/calc_qc_fc_correlations.R")
FDFC=calc_qc_fc(final)
write.csv(FDFC,"report/FDFCcorrelations.csv", row.names = FALSE)
```

```{r figNetworks, echo=FALSE, out.width="49%", out.height="30%", fig.cap="Resting state networks based on AROMA+CC+GSR input data, one-sample t-tests adjusting for age and sex over all timepoints and participants with bootstrapped clusterwise inference (p FWE<0.05)", fig.subcap=c("DMN seeded from precuneus", "Reward network seeded from Nucleus Accumbens"),fig.show='hold',fig.align='center'}
#setwd(knitr::opts_knit$get("root.dir"))
networks = c("/data/pt_02161/Analysis/Project2_resting_state/seed-based/Second_level /code_and_manuscript/report/DMN_thresholded_for_clusterFWE005.pdf",
             "/data/pt_02161/Analysis/Project2_resting_state/seed-based/Second_level /code_and_manuscript/report/Rew_thresholded_for_clusterFWE005.pdf")
knitr::include_graphics(networks)
```

@@

```{r prepare tsnr ROI plot, echo=FALSE}
fig_tSNR <- mk_figtSNR(final)
```

```{r figtSNR,fig.cap="Lower tSNR in the region of interest NAcc versus precuneus used for seed-based connectivity (shown are all 101 timepoints)", out.width="60%", fig.align="center", echo=FALSE}

fig_tSNR
```
Yet, the reward network was less pronounced (overall lower T-values) and symmetric (no significant connectivity with left amygdala), which might be due to lower SNR in the seeded brain region (see Figure \@ref{fig:figtSNR}.

@@

### DMN/reward network with GM mask (exploratory analysis)
<!-- GM mask: Group*Time Interaction [OLD]-->
<!-- Parametric and non-parametric significance test for the model testing for time-by-group interaction did not show any different results when using a grey matter mask. Again, parametric estimation and non-parametric estimation differed in regard to the time-by-group interaction of PCC and voxel ($18,42,30$) in the connectivity maps without GSR correction. While co-activation was significant in parametric estimation ($Z = 4.72$, $q_{FDRcorr} = 0.040$), it was non-significant when using wild bootstrap ($q_{\mathrm{FDRcorr}} = 0.998$)<!-- also FWE-corr n.s -->

## Interaction effect group by time on aggregated FC (preregistered analysis)
```{r load agg FC data and prepare, echo=FALSE}
agg_FC_DMN=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/DMN/agg_FC_DMN.txt")
agg_FC_DMN_ID=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/DMN/agg_FC_DMN_ID.txt")
agg_FC_Rew_ID=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/Rew/agg_FC_Rew_ID.txt")
agg_FC_Rew=read.table("/data/pt_02161/Results/Project2_resting_state/connectivity/Analysis/aggFC/Rew/agg_FC_Rew.txt")
for (i in c(1:nrow(agg_FC_DMN_ID))){
  agg_FC_Rew[i,"subj.ID_tp"]=strsplit(base::strsplit(as.character(agg_FC_Rew_ID[i,"V1"]),'/')[[1]][10],'_nacc')[[1]][1]
  agg_FC_DMN[i,"subj.ID_tp"]=strsplit(base::strsplit(as.character(agg_FC_DMN_ID[i,"V1"]),'/')[[1]][10],'_pcc')[[1]][1]
}

>>>>

---
title: "Effects of bariatric surgery and weight loss on resting state functional connectivity"
author:
- Hannah Sophie Heinrichs* [1]
- Frauke Beyer* [1][2]
- Kerstin Prehn
- Agnes Floel
- Arno Villringer [1][2][3]
- A. Veronica Witte [1][2][3]
date: 1 Max-Planck-Institute for Human Cognitive and Brain Sciences, Leipzig \newline
  2 CRC 1052 "Obesity Mechanisms", Subproject A1, University of Leipzig \newline 3
  Day Clinic for Cognitive Neurology, University Clinic Leipzig
output:
  bookdown::pdf_book:
    fig_caption: yes
    always_allow_html: true
    number_sections: yes
    toc: yes
    extra_dependencies: "subfig"
bibliography: "bibliography.bib"
editor_options:
  chunk_output_type: console

header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

```{r "knit setup", include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE) # fig.height=2, fig.width=4)
options(knitr.table.format = "latex")
options(knitr.kable.NA = '') # hide NA in tables
# set working directory of knit (RMarkdown) to directory of Rproj
#opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
#setwd(knitr::opts_knit$get("root.dir"))

@@

```{r "Load data", include=FALSE}
#final <- create_sample_df(group = "both", tp = "all")
final <- read.csv(file = "../report/final.csv")

@@

```{r "tableSample", echo=FALSE}
#tableSample <- mk_SampleTable(final)
tableSample <- readRDS(file = "../report/tab/tableSample.rds")

@@

<!-- STUDY DESIGN --> The ADIPOSITAS-Study conducted in the Charité Berlin is a observational, prospective and longitudinal design described in detail by @Prehn_2020. Data collection is still ongoing but only data available in April 2019 is used.
The procedures followed were in accordance with the Helsinki Declaration. This study was registered at clinicaltrials.gov as NCT01554228 and described in more detailed in the preregistration under [https://osf.io/yp42s](https://osf.io/yp42s). <!-- previous version: https://osf.io/f8tpn, update: https://osf.io/59bh7/ -->
```{r tableDescr, echo=FALSE}
# table for descriptive statistics
tableDescr <- readRDS(file = "../report/tab/tableDescr.rds")

knitr::kable(
    tableDescr,
    row.names = FALSE,
    format = "latex",
    booktabs = T,
    linesep = "",
    caption = "BMI and mean FD of participants acquired in each condition and time point")
```
<!-- RECRUITEMENT, INCLUSION/EXCLUSION CRITERIA --> Participants were recruited from the Center for Bariatric and Metabolic Surgery if they were between age $18$ and $70$ and fulfilled the inclusion criteria. These were, in accordance with BARS guidelines, a failure of conservative obesity treatment and either (1) a BMI $> 40$ km/m$^2$ or (2) a BMI $> 35$ kg/m$^2$ and at least one typical co-morbidity (e.g. type-2 diabetes, hypertension, non-alcoholic fatty liver disease)[@Mechanick_2013].
Participants with a history of cancer, chronic inflammatory disease and addiction, other severe untreated diseases, brain pathologies identified in the MRI scan or cognitive impairments (MMSE score $< 34$) were excluded from the study.
Measures were taken at baseline (BL), $6$ (FU1) and $12$ (FU2) months post-surgery or after the baseline appointment but not all participants have complete data for all time points (see Table \@ref(tab:tableDescr)). `r sum(freqtable$Freq == 3)` participants had complete data, `r sum(freqtable$Freq == 2)` provided data for two time points and `r sum(freqtable$Freq == 1)` for only one time point.
The current sample is the subsample of $69$ subjects that were originally enrolled in the study entailing only subjects that have received MRI. Analyses were performed on all participants that provided at least one data point of fMRI data (detailed information on data availability in \@ref(tab:tableSample)).

<!-- DESCRIPTIVE STATISTICS --> The final sample entailed morbidly obese individuals 
(n = `r nrow(subj_df)`, 
`r nrow(filter(subj_df,Sex == "f"))` female; 
`r round(mean(subj_df$Age_BL),2)` 
$\pm$ `r round(sd(subj_df$Age_BL),2)` SD years old
(range `r min(subj_df$Age_BL)`-`r max(subj_df$Age_BL)`), 
that either underwent surgery 
(n = `r cs["Sum","IG"]`, 
`r cs["f","IG"]` female, 
`r round(mean(subj_df$Age_BL[subj_df$condition == "IG"]),2)` 
$\pm$ `r round(sd(subj_df$Age_BL[subj_df$condition == "IG"]),2)` SD years old) 
or were waiting list controls 
(n = `r cs["Sum","KG"]`, 
`r cs["f","KG"]` female,
`r round(mean(subj_df$Age_BL[subj_df$condition == "KG"]),2)`
$\pm$ `r round(sd(subj_df$Age_BL[subj_df$condition == "KG"]),2)` SD years old) 
The study design was quasi-experimental, as patient in the waiting control list had also been recommended to undergo bariatric surgery, but were still waiting for their health insurance's approval.
```{r tableBaselineCharacteristics, include=FALSE}
#dfBL <- create_sample_df(group = "all", tp = "BL")
dfBL <- readRDS(file = "../report/dfBL.rds")

@@

# Sex
p <- ggplot(data = dfBL, aes(x = condition, fill = Sex)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(labels = c("female", "male"),
                    values = c("#00305E55", "#C0004555")
  )
tab_sex <- table(dfBL$condition, dfBL$Sex)
```

```{r figBMIdescr, fig.align='center', out.width="90%", fig.cap="BMI over time.", echo=FALSE}
# eval = TRUE, fig.asp = 0.7, fig.width = 6
readRDS(file = "../report/fig/fig_BMIdescr.rds")
```
<!-- BASELINE CHARACTERISTICS --> Histograms on baseline characteristics revealed, that patients in the control group did not differ notably from the intervention group regarding BMI and log mean FD, and only slightly in distribution of sex and age, the control group had a higher number of male participants (n = `r tab_sex[2,2]` vs. n = `r tab_sex[2,1]`) and a higher mean age (`r round(tab_age$mean[tab_age$group1=="KG"],2)` vs. `r round(tab_age$mean[tab_age$group1=="IG"],2)`). Change in BMI throughout the study is depicted in \@ref(fig:figBMIdescr).
<!-- SURGICAL PROCEDURE --> As intervention, $0$ participants received surgery that combines restrictive and malabsorptive mechanisms (RYGB) and $0$ surgery purely restricting energy intake (VSG: n = $0$, GB: n = $0$).
<!-- STUDY PROCEDURE --> Subjects arrived in the morning (between 07:00 and 12:00) after an overnight fast. They underwent medical assessments including an interview, blood draw and anthropocentric measurements before having a one our break for <!-- standadized??  -->breakfast. MRI scanning was done after further psychological tests (for details see @Prehn_2020).

@@

### Minimal functional preprocessing
The function preprocessing module included the removal of first four volumes, motion correction (FSL's `MCFLIRT`), fieldmap distortion correction (FSL's `fsl_prepare_fieldmap` and `FUGUE`) and coregistration to the subject’s individual longitudinal anatomical space (FreeSurfer's `bbregister`). In more detail, the transformations derived from the latter three steps were combined into one and applied in a single step.
For further analysis and ICA-AROMA processing, the minimally preprocessed data were intensity normalized and smoothed with a 6mm Gaussian kernel (`fslmaths -kernel gauss 2.548`).
<!-- *This part of the analysis was described in the preregistration as "After removing the first four volumes, transforms for motion correction (FSL’s MCFLIRT), fieldmap distortion correction (based on fieldmap and FSL’s FUGUE) and coregistration to the subject’s individual anatomical space (using FREESURFER’s bbregister) are calculated."*   -->

@@

### Functional connectivity
To derive reward and DMN functional connectivity maps, we used Nucleus Accumbens (Nacc) and precuneus as seed regions of interest (ROI), respectively. Based on FreeSurfer's `aseg.mgz` and Desikan-Killiany parcellation, we created seed masks using `mri_binarize` (thresholded for NAcc at 26,58; precuneus at 1025,2025) and averaged them over hemispheres. Then, we used `NiftiLabelsMasker` and `NiftiMasker` to extract the standardized timeseries from the seed regions and the whole brain. We calculated the Pearson's correlation between them with `numpy.dot`, performed r-to-z transform and saved the resulting correlation maps for each processing mode (minimally preprocessed, aroma+cc, aroma+gsr). Finally, the connectivity maps were transformed into MNI space using the affine transformation and non-linear warp derived with ANTS during anatomical preprocessing.
<!-- *In the preregistration we stated "Resting-state functional connectivity (FC) using seed-based connectivity analysis extracted with in-house pre-processing pipelines (see (Beyer et al., 2019)): Whole-brain FC maps for a priori defined seed regions (defined based on the results of the subcortical segmentation and Desikan-Killiany cortical parcellation in the Freesurfer version 6.0.0 longitudinal processing stream). For the reward network a ROI combined from bilateral Nucleus accumbens from the subcortical segmentation (26, Left-Accumbens-area, 58, Right-Accumbens-area) was selected. Ventro-medialprefrontal cortex was not selected as seed region due to low signal-to-noise ratio in this part of the brain in many participants (noticed during quality checking). For the default mode network, the seed region was combined from bilateral precuneus from the Desikan-Killiany atlas (1025 ctx-lh-precuneus, 2025 ctx-rh-precuneus). We will calculate  voxel-wise seed-based connectivity mapsin the individual subject’s space (which is coregistered to the Freesurfer longitudinally preprocessed timepoint) and apply r-to-Z-transform to these images. These steps will be performed using nilearn, nipype and python. Then, we will transform the connectivity maps into MNI space with a non-linear warp derived from anatomical preprocessing of the longitudinally preprocessed timepoints (using ANTS).* -->

@@

FreeSurfer cross-sectional and longitudinal  segmentations were visually checked according to @Klapwijk_2019. We excluded 5 datasets from 3 participants because of excessive head motion leading to failed pial reconstruction and anatomical-functional coregistration .
<!--ADI009_fu2, ADI063_bl, ADI063_fuADI116_bl, ADI116_fu, ADI116_fu2-->
<!-- *In the preregistration we stated that "Freesurfer segmentations and surfaces will be visually controlled and corrected according to published recommendations, if necessary", and that "We will exclude participants in which head motion artifacts affected the T1-weighted image preprocessing (e.g. caused Freesurfer to fail or perform very poorly (rated “Failed” according to (Klapwijk et al., 2019)".* -->

@@

<!-- *In the preregistration we wrote: "We will use the average of connectivity values within masks of the respective networks as aggregate network connectivity measure. The mask will include all voxels surviving FWE-correction of p<0.05 in a one-sample t-test of the individual connectivity maps across all participants and time points, calculated with the modified SwE option in the SwE toolbox"* -->

@@

DesignMatricesList <- readRDS(file = "../report/fig/DesignMatricesList.rds")
DesignMatricesList$plotA
DesignMatricesList$plotB
#gridExtra::grid.arrange(plotA, plotB, nrow=1)
```
For the second and third model, we calculated the centered mean BMI and log mean FD per subject and the intra-subject-centered BMI as proposed by @Guillaume_2014. With a model containing average BMI and longitudinal change per subject, we investigated their differential effects on the complete sample and subsequently in a exploratory fashion for the intervention group only as the effect of longitudinal change in BMI was expected to be driven by this subsample.
<!-- NUISANCE COVARIATES --> Baseline age, sex, and log mean framewise displacement (FD) were entered into the first and second model as nuisance variables. Age and sex were between-subject covariates while FD is time-varying. Mean FD was log-transformed and age was demeaned by subtracting the overall mean across subjects and time points.
As change in FD and change in BMI share common variance [@Beyer_2019], the inclusion of log mean FD may be disputed and done optionally. For that reason, we performed each analysis separately one with mean FD (three and without. Differences in results will be reported. We consider this sensitivity analysis theoretically more meaningful than the preregistered comparison with the nuisance covariate free model and a model with four nuisance covariates (including baseline BMI).<!-- Prereg: We will perform sensitivity analysis adjusting for baseline age, sex, average of mean FD of both time points and baseline BMI -->
<!-- POST HOC  --> For a better understanding of the unique contribution of average and longitudinal change FD measures, we separately employed a third post-hoc model to disentangle the effects: $FC = between-subject FD + within-subject FD$ with nuisance covariates age and sex. As these post-hoc analyses were not preregistered, they should be considered to be exploratory.

@@

## Quality control
We performed different quality controls for the rsfMRI data.
As described in @Beyer_2020, group and time show an interaction effect for mFD where the intervention group reduced in head motion in the follow-up assessments compared to the control group.
We also inspected DVARS, a measure of whole-brain signal variation which is reflective of noise in the data. We saw that DVARS did not qualitatively differ between groups and time points, and that the correlation of DVARS and mFD remained similar over time points.
This indicates that even though there was a decrease in head motion in the intervention versus control group, the impact of head motion on imaging data remained similar over the study period.

```{r mFDFC, echo=FALSE}
res <- read.csv("../report/FDFCcorrelations.csv")
```

```{r tableQcfc, echo=FALSE}
tab_Qcfc <-
  knitr::kable(
    round(res, 3),
    col.names = colnames(res),
    row.names = TRUE,
    format = "latex",
    booktabs = T,
    linesep = "",
    caption = "Summary of quality metrics for different denoising pipelines across conditions and time points"
  ) %>%
  column_spec(c(2:6), width = "1.8cm")
tab_Qcfc
```

Further, we evaluated the impact of our denoising pipelines on different quality metrics following [@Ciric_2018]. Table \@ref(tab:tableQcfc) shows that in the minimally preprocessed data, mFD and FC between nodes was significantly positively correlated (mean Spearman's r = `r round(res[1,1],2)`). This relationship was negatively associated with distance between nodes (mean Spearman's r = `r round(res[1,5],2)`).  
Denoising procedures (AROMA, AROMA+CC, AROMA+CC+GSR) gradually reduced the correlation of mFD and connectivity, and reduced the distance dependency. Contrary to previous reports, GSR did not exercabate mFD-FC distance dependency. Overall, AROMA+CC+GSR performed best, with almost no mFD-FC correlation (mean Spearman's r = `r round(res[4,1],2)`), and minimal positive distance dependence (mean Spearman's r = `r round(res[4,5],2)`). Yet, AROMA+CC also performed well, and we thus calculated our analysis models for both preprocessing schemes.

```{r figDvarsmFD, fig.align = "center", out.width="100%", fig.cap='Summary of quality measures DVARS and mFD', echo=FALSE}

figList <- readRDS(file = "../report/fig/figDvarsmFDList.rds")
(figList$fig1 | figList$fig2) / figList$fig3
```

```{r figNetworks, echo=FALSE, out.width="49%", out.height="30%", fig.cap="Resting state networks based on AROMA+CC+GSR input data, one-sample t-tests adjusting for age and sex over all timepoints and participants with bootstrapped clusterwise inference (p FWE<0.05)", fig.subcap=c("DMN seeded from precuneus", "Reward network seeded from Nucleus Accumbens"),fig.show='hold',fig.align='center'}
#setwd(knitr::opts_knit$get("root.dir"))
networks = c("../report/fig/DMN_thresholded_for_clusterFWE005.pdf",
             "../report/fig/Rew_thresholded_for_clusterFWE005.pdf")
knitr::include_graphics(networks)
```

@@

```{r figtSNR,fig.cap="Lower tSNR in the region of interest NAcc versus precuneus used for seed-based connectivity (shown are all 101 timepoints)", out.width="60%", fig.align="center", echo=FALSE}
readRDS(file = "../report/fig/fig_tSNR.rds")
```
Yet, the reward network was less pronounced (overall lower T-values) and symmetric (no significant connectivity with left amygdala), which might be due to lower SNR in the seeded brain region (see Figure \@ref{fig:figtSNR}.

@@

## Interaction effect group by time on aggregated FC (preregistered analysis)
```{r load agg FC data and prepare, echo=FALSE}
final_FC <- read.csv("../report/final_FC.csv")
```
